A free/open-source machine translation system for English to Kazakh

### 2015-03-02 engkaz-eamt-2015/engkaz-eamt-2015.tex
more
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59188 72bbbca6-d526-0410-a7d9-f06f51895060

\section{The Apertium platform}
Apertium \citep{forcada11} is a free/open-source rule-based machine translation (MT) platform that 
has been developed since 2005 at the Universitat d'Alacant. At first, it was initially aimed at 
translating texts between closely related languages, then it was extended to deal with less-related 
languages. This platform has next components: machine translation engine, developer's tools, and linguistic 
data for an increasing number of language pairs and they are licensed under the 
free/open-source GNU General Public License (GPL, versions 2 and 3).

@@

@@

  and inflection information. For example, the morphological analyser will deliver the following analysis for 
  the sentence ``I played''.
\begin{verbatim}
^I/I<prn><subj><p1><mf><sg>$ 
^played/play<vblex><past>/play<vblex><pp>$
\end{verbatim}
  In this example each source form has been analysed as one or more lexical forms: ``I'' is analysed 
  into ``I'', where lexical category is subject pronoun (\texttt{<prn><subj>}) and first person (\texttt{<p1>}), could be 
  masculine or feminine (\texttt{<mf>}), singular (\texttt{<sg>}); ``played'' is analysed into lexical verb (\texttt{<vblex>}) 
  ``play'' with past simple tense (\texttt{<past>}) or it could be past participle (\texttt{<pp>}), so it has two 
  analyses. Each analysis of word is separated by \texttt{\^{}} and \texttt{\$} symbols, and for one word each lexical 
  forms are delimited by ``/'', and tags(<...>) show grammatical attributes of lexical form. 

@@

  analyses using rules based on context. For example, consider the morphological analysis of word ``book'':
\begin{verbatim}
^book/book<n><sg>
/book<vblex><inf>
/book<vblex><pres>$
\end{verbatim}
  This word has 3 lexical forms, and depends on context, ont of them will be chosen by rules or tagger. After  this module, 
  all words have only one morphological analysis.

@@

  lexical form. This module uses a bilingual dictionary. For the English-Kazakh language pair, each lexical form of 
  English word is translated into Kazakh as follows:
\begin{verbatim}
^I<prn><subj><p1><mf><sg>/мен<prn><pers><subj><p1><mf><sg>$ 
^play<vblex><past>/ойна<v><tv><past>$
\end{verbatim}

@@

  lexical changes such as changes in prepositions and agreement between target language lexical forms. Transfer rules can 
  produce new sequences for the  target language, for instance, preposition-noun rule is used to built sequence, where for 
  noun is chosen right case, which depends on preposition: 
\begin{verbatim}
in garden  --  ^бақша<n><sg><PXD><loc>$
 \end{verbatim} 

@@

Additional cleaning stage was created for removing extra tags which did not determine through the chunker and 
interchunk steps, for instance, noun was not acting as object and did not get accusative case, so not determined 
tag was \texttt{<CD>} after ``cleaning'' it will be set on nominative case; at this step, also carried out 
agreement between noun or adjective and copula ``е''(Мен дәрігер+е<p1 person singualr>), vowel and consonants 
agreement between last word of question and question particle (Сіз келдіңіз + ме? - Сіз келдіңіз + бе?).

@@

Evaluating the output of the system was measured with BLEU \citep{papineni02} and Word Error Rate 
metric \citep{levenshtein/1966}. Was chosen text small English text, and by postediting output 
of English-Kazakh system, was build parallel text.

@@

We have presented the rule-based MT system from English to Kazakh. The current English-Kazakh machine translation system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and contains a good vocabulary for testing purposes. 
We plan to continue developing English-Kazakh pair, going to extend the coverage to 98\% on the corpora, which we used before. Additionally, we will improve the quality of translation by adding more rules, such like constraint grammar rules, structural transfer and lexical rules. The future plan is to use the created data with other open-source MT system with Turkic languages or system with Kazakh language on right side to make transfer systems between the Turkic or other language pairs. Related work is currently ongoing with Russian--Kazakh. 
Our system is available as free/open-source software and the whole system may be downloaded from sourceforge(apertium-eng-kaz1, apertium-kaz2).

@@

%\section{References}
%1 Сундетова А.М., Кәрібаева А.С., АПЕРТИУМ ПЛАТФОРМАСЫНДАҒЫ АҒЫЛШЫН-ҚАЗАҚ МАШИНАЛЫҚ АУДАРМАШЫ ҮШІН ЕКІТІЛДІ СӨЗДІКТІ ҚҰРУ. Матер. межд. научн.-практ. конф. «Применение информационно-коммуникационных технологий в образовании и науке», посвященной 50-летию Департамента информационно-коммуникационных технологий и 40-летию кафедры «Информационные системы» КазНУ им. аль-Фараби. 22 ноября 2013г. – Алматы: Қазақ Университеті, 2013. – С.53-57.

>>>>

\section{The Apertium platform}

@@

Apertium \citep{forcada11} is a free/open-source rule-based machine translation (MT) platform that 
has been developed since 2005 at the Universitat d'Alacant. At first, it was initially aimed at 
translating texts between closely related languages, then it was extended to deal with less-related 
languages. This platform has next components: machine translation engine, developer's tools, and linguistic 
data for an increasing number of language pairs and they are licensed under the 
free/open-source GNU General Public License (GPL, versions 2 and 3).

@@

  and inflection information. 
%  In this example each source form has been analysed as one or more lexical forms: ``I'' is analysed 
%  into ``I'', where lexical category is subject pronoun (\texttt{<prn><subj>}) and first person (\texttt{<p1>}), could be 
%  masculine or feminine (\texttt{<mf>}), singular (\texttt{<sg>}); ``played'' is analysed into lexical verb (\texttt{<vblex>}) 
%  ``play'' with past simple tense (\texttt{<past>}) or it could be past participle (\texttt{<pp>}), so it has two 
%  analyses. Each analysis of word is separated by \texttt{\^{}} and \texttt{\$} symbols, and for one word each lexical 
%  forms are delimited by ``/'', and tags(<...>) show grammatical attributes of lexical form. 

@@

  analyses using rules based on context. 
%For example, consider the morphological analysis of word ``book'':
%\begin{verbatim}
%^book/book<n><sg>
%/book<vblex><inf>
%/book<vblex><pres>$
%\end{verbatim}
%  This word has 3 lexical forms, and depends on context, ont of them will be chosen by rules or tagger. 
After  this module, all words have only one morphological analysis.

@@

  lexical form. This module uses a bilingual dictionary. 
%For the English-Kazakh language pair, each lexical form of 
%  English word is translated into Kazakh as follows:
%\begin{verbatim}
%^I<prn><subj><p1><mf><sg>/мен<prn><pers><subj><p1><mf><sg>$ 
%^play<vblex><past>/ойна<v><tv><past>$
%\end{verbatim}

@@

  lexical changes such as changes in prepositions and agreement between target language lexical forms. %Transfer rules can 
%  produce new sequences for the  target language, for instance, preposition-noun rule is used to build a sequence, where for 
%  noun is chosen right case, which depends on preposition.
%\begin{verbatim}
%in garden  --  ^бақша<n><sg><PXD><loc>$
% \end{verbatim} 

@@

Additional cleaning stage was created for removing extra tags which did not determine through the chunker and 
interchunk steps, for instance, noun was not acting as object and did not get accusative case, so not determined 
tag was \texttt{<CD>} after ``cleaning'' it will be set on nominative case; at this step, also carried out 
agreement between noun or adjective and copula ``е''(Мен дәрігер+е<p1 person singular>), vowel and consonants 
agreement between last word of question and question particle (Сіз келдіңіз + ме? - Сіз келдіңіз + бе?).

@@

Evaluating the output of the system was measured with BLEU \citep{papineni02} and Word Error Rate 
metric \citep{levenshtein/1966}. We chose a short text in English, and by postediting output 
of English-Kazakh system, was build parallel text. The output of each of the machine
translation systems was posteditted independently to avoid biasing one particular system.

@@

We have presented the rule-based MT system from English to Kazakh. The current English-Kazakh machine translation 
system already successfully translates noun-phrases, verb-phrases, prepositional-phrases, and adjectival-phrases, and 
contains a good vocabulary for testing purposes. 
We plan to continue developing English-Kazakh pair, going to extend the coverage to 98\% on the corpora, 
which we used before. Additionally, we will improve the quality of translation by adding more rules, such like 
constraint grammar rules, structural transfer and lexical rules. The future plan is to use the created data with other 
open-source MT system with Turkic languages or system with Kazakh language on right side to make transfer systems 
between the Turkic or other language pairs. Related work is currently ongoing with Russian--Kazakh. 
Our system is available as free/open-source software and the whole system may be downloaded 
from SourceForge.\footnote{\url{https://svn.code.sf.net/p/apertium/svn/incubator/apertium-eng-kaz}}

@@
